The brightness and spatial distributions of terrestrial radio sources

### 2013-01-17 articles/Article-RFI-distribution/article-rfi-distribution.tex
Fixed several formulae and fixed all ref I comments

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
% (c=rho)
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by scaling and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, one needs to assume that there are no samples beyond some limiting distance $r_L$ to invert the cumulative function. With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
S_{\textrm{RFI}} \leftarrow \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $x_u, y_u \sim U(0, 1)$. The result is an amplitude sample $S$, with $S \ge Ig/r_L^2$, that follows a power-law distribution. The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
\textrm{real}(S) \leftarrow x_n + \frac{Ig \cos 2 \pi y_u}{x_u r_L^2}
\textrm{,\hspace{1mm} imag}(S) \leftarrow y_n + \frac{Ig \sin 2 \pi y_u}{x_u r_L^2},
\end{equation}
with $x_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}({S})^2+\textrm{imag}({S})^2}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

For completeness, we also show the formula for sampling apparent source strengths that include surface propagation effects, which can be used for sampling realistic source amplitudes in simulations. This is calculated by inversion and scaling of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation}
\textrm{real}(S_{\textrm{RFI}}) \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta},
\textrm{imag}(S_{\textrm{RFI}}) \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta}.
\end{equation}

@@

We assume that the stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, also has little effect on the $\log N$ -- $\log S$ histograms. Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls). Therefore, $I$ might follow a power law by itself. On average, strong transmitters transmit longer, thus will also contaminate more samples. High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW. Low-power transmitters, such as remote controls, transmit with an order of 100 mW. Therefore, these devices have a spread of around 6 orders of magnitudes in power. As long as the exponent of this distribution is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes. With a spatial $-1.5$ power law and the given transmitting powers, the low-power transmitters should contaminate a factor of $10^9$ more samples compared to the high-power transmitters to dominate the high-amplitude distribution, which is unlikely.

@@

One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator $\hat \alpha_H$ \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \hat \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{min}} \right)^{-1},
\end{equation}
with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples drawn from a power law with lower bound $x_\textrm{min}$.
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution. Moreover, the estimator requires reading and processing all data points, and because of the size of the data sets this is somewhat impractical. However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency. Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator. This avoids the need to iterate over all data.

@@

In experiments such as the LOFAR EoR project, a simulation pipeline is used to create a realistic estimate of the signal that can be expected. Currently, these simulations do not include the effects of RFI. With the construction of empirical models for the RFI source distributions, we are one step closer to including these effects in the simulation. Using Eq.~\eqref{eq:drawing-rfi-distribution}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger. What is still needed for accurate simulation, is to obtain a likely distribution for the duration that one such source affects the data. For example, it is neither realistic that all RFI sources are continuously transmitting nor that they affect only one sample. The RFI detector is highly depending on the morphology of the feature in the time-frequency domain. Finally, the coherency properties of the RFI might be even more important to simulate correctly, but these have been not been explored. However, these have large implications for observations with high sensitivity. This will be discussed in the next section.

@@

\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution, before further averaging. There are several reasons for this, which we will summarize one by one.
\begin{itemize}
 \item Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes. This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
 \item Many RFI sources will be seen by only a few stations. This would make the histogram go down more quickly at lower amplitudes.
 \item For the shortest baselines at 150~MHz, the far field starts around 1~km. Some RFI sources will therefore be in the near field, especially for the longer baselines. In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.
 \item Many stationary sources are not constant over time, thus will be attenuated somewhat by averaging.
 \item We have assumed 100\% of the spectrum is occupied by RFI. If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
 \item Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other. Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread. Therefore, they will add incoherently.
\end{itemize}

Two other effects can improve the RFI situation as well:
\begin{itemize}
 \item Fringe stopping interferometers can partly average out RFI sources. Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}. This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
 \item The Fourier transform that is involved in data imaging will localize the contribution from stationary RFI near the NCP. If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.
\end{itemize}

@@

>>>>

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
% (c=rho)
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by scaling and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, one needs to assume that there are no samples beyond some limiting distance $r_L$ to invert the cumulative function. With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
S_{\textrm{RFI}} \leftarrow \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $x_u, y_u \sim U(0, 1)$. The result is an amplitude sample $S$, with $S \ge Ig/r_L^2$, that follows a power-law distribution. The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
S \leftarrow v_n + w_n i + \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}^2({S})+\textrm{imag}^2({S})}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

For completeness we show the formula for sampling apparent source strengths that include surface propagation effects, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 \le x_u,y_u < 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.

@@

We assume that the stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, also has little effect on the $\log N$ -- $\log S$ histograms. Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls). Therefore, variable $I$ might follow a power law by itself. It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well. High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW. Low-power transmitters, such as remote controls, transmit with an order of 100 mW. Therefore, these devices have a spread of around 6 orders of magnitudes in power. As long as the exponent of the power law of $I$ is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes. With a spatial $-1.5$ power law and the given transmitting powers, the low-power transmitters should contaminate a factor of $10^9$ more samples compared to the high-power transmitters to dominate the high-amplitude distribution, which is unlikely. Therefore, it is likely that the spatial distribution will dominate the power law in the histogram.

@@

One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator $\hat \alpha_H$ \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \hat \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{min}} \right)^{-1},
\end{equation}
with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with lower bound $x_\textrm{min}$.
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution. Moreover, the estimator requires reading and processing all data points, and because of the size of the data sets this is somewhat impractical. However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ is the bin frequency. Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator. This avoids the need to iterate over all data.

@@

In experiments such as the LOFAR EoR project, a simulation pipeline is used to create a realistic estimate of the signal that can be expected. Currently, these simulations do not include the effects of RFI. With the construction of empirical models for the RFI source distributions, we are one step closer to including these effects in the simulation. Using Eq.~\eqref{eq:sample-with-propagation}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger. What is still needed for accurate simulation, is to obtain a likely distribution for the duration that one such source affects the data. For example, it is neither realistic that all RFI sources are continuously transmitting nor that they affect only one sample. The RFI detector is highly depending on the morphology of the feature in the time-frequency domain. Finally, the coherency properties of the RFI might be even more important to simulate correctly, but these have been not been explored. However, these have large implications for observations with high sensitivity. This will be discussed in the next section.

@@

\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution, before further averaging. There are several reasons for this, which we will summarize one by one.

There are several effects related to the origin of the interference, which can reduce their effect. One such effect, is when the RFI sources have a variable geometric phase. Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes. This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise. Moreover, many RFI sources will be seen by only a few stations or are not constant over time, which further attenuates their effect. Finally, for the shortest baselines at 150~MHz, the far field starts around 1~km. Some RFI sources will therefore be in the near field, especially for the longer baselines. In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see it with different phases.

A large number of RFI sources together can also behave like normal incoherent noise. Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other. Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread. Therefore, they will add incoherently.

Sometimes, standard data processing also attenuates RFI, mainly because of fringe stopping and the imaging of data. Fringe stopping interferometers can partly average out RFI sources. Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}. This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP. Imaging sometimes attenuates RFI, because the Fourier transform that is involved in data imaging will localize the contribution from stationary RFI near the NCP. If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.

@@

On top of all these possible RFI reducing effects, our given RFI constraints could be too pessimistic because we have assumed 100\% of the spectrum is occupied by RFI. If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
